Investigating Biases Associated with Dietary Starch Incorporation and Retention with an Oral Biofilm Model

### 2022-05-31 analysis/paper/abstract.Rmd
incorporate reviewer comments

Dental calculus has proven to contain a wealth of information on the
dietary habits of past populations. These insights have, to a large
extent, been obtained by the extraction and identification of starch
granules contained within the mineralised dental plaque from a wide
range of regions and time periods. The scope of previous studies have
been limited to microfossil extraction and identification to reconstruct
dietary preferences from the archaeological record, and few studies have
attempted to address the biases of starch retention in dental calculus.
Those that have considered this problem have been limited to \emph{in
vivo} studies on modern humans and non-human primates. Here, we present
a multispecies oral biofilm model, which allows experimental research on
starch incorporation and retention to be conducted on \emph{in vitro}
dental calculus in a controlled laboratory setting. The biofilms were
exposed to treatment solutions with known quantities of dietary starches
(wheat and potato) during the 25-day growth period. After this, the
starch granules were extracted from the mature biofilm (by dissolution
in EDTA), and counted. We show that the granule counts extracted from
the model dental calculus represented a low proportion (ranging from
0.06\% to 0.16\%) of the total number of granules exposed to the
biofilms throughout the experiment. Additionally, we found that the
ratios of granule sizes from the extracted starch granules differed from
the original treatment solutions, with large granules (\textgreater20
\(\mu\)m) consistently being under-represented. We also found a
correlation between the absolute granule counts and dry-weight of the
biofilm (\emph{r} = 0.659, 90\%CI{[}0.463, 0.794{]}), as well as between
the concentration (count per mg) of granules and dry-weight (\emph{r} =
0.3, 90\%CI{[}0.0618, 0.506{]}).\\
Our results reinforce previous \emph{in vivo} studies suggesting that
dental calculus presents a very small, and partly biased picture of the
original dietary intake of starches, with an over-representation of
plants producing granules smaller than 20 \(\mu\)m in size. The
experimental model presented here is well-suited to address the need for
further validation of methods and biases associated with dietary
research on dental calculus.

>>>>

Dental calculus has proven to contain a wealth of information on the
dietary habits of past populations. These insights have, to a large
extent, been obtained by the extraction and identification of starch
granules contained within the mineralised dental plaque from a wide
range of regions and time periods. The scope of previous studies have
been limited to microfossil extraction and identification to reconstruct
dietary preferences from the archaeological record, and few studies have
attempted to address the biases of starch retention in dental calculus.
Those that have considered this problem have been limited to \emph{in
vivo} studies on modern humans and non-human primates. Here, we present
a multispecies oral biofilm model, which allows experimental research on
starch incorporation and retention to be conducted on \emph{in vitro}
dental calculus in a controlled laboratory setting. The biofilms were
exposed to treatment solutions with known quantities of dietary starches
(wheat and potato) during the 25-day growth period. After this, the
starch granules were extracted from the mature biofilm (by dissolution
in EDTA), and counted. We show that the granule counts extracted from
the model dental calculus represented a low proportion (ranging from
0.06\% to 0.16\%) of the total number of granules exposed to the
biofilms throughout the experiment. Additionally, we found that the
ratios of granule sizes from the extracted starch granules differed from
the original treatment solutions, with large granules (\textgreater20
\(\mu\)m) consistently being under-represented. We also found a
positive correlation between the absolute granule counts and dry-weight of the
biofilm (\emph{r} = 0.659, 90\%CI{[}0.463, 0.794{]}), meaning the absolute
quantity of starch granules will increase as the size of the calculus deposit
increases. A similar, but weaker correlation was found between
the concentration (count per mg) of granules and dry-weight (\emph{r} =
0.3, 90\%CI{[}0.0618, 0.506{]}).\\
Our results reinforce previous \emph{in vivo} studies suggesting that
dental calculus presents a very small, and partly biased picture of the
original dietary intake of starches, with an over-representation of
plants producing granules smaller than 20 \(\mu\)m in size. The
experimental model presented here is well-suited to address the need for
further validation of methods and biases associated with dietary
research on dental calculus.

### 2022-05-31 analysis/paper/discussion.Rmd
incorporate reviewer comments

We have also shown that the size of the starch granules influences the likelihood
of incorporation into the calculus. Starch granules larger than 20 $\mu$m in
maximum length were underrepresented in the calculus samples compared to the original 
starch solutions, an effect that was consistent across all three treatments. Medium
granules (10--20 $\mu$m) were often over-represented
(Table \@ref(tab:sample-prop-tbl), 
and 
Figure \@ref(fig:ratio-plots)).
Large potato granules were most affected, potentially because of the greater
size-range. They can reach up to 100 $\mu$m in maximum length, whereas wheat 
granules generally only reach up to 35 $\mu$m 
[@gismondiStarchGranulesData2019; @haslamDecompositionStarch2004; @seidemannStarchAtlas1966, 174-176].
Granule morphology may also play a role. Large wheat granules 
are lenticular and have a larger surface area compared 
to volume, whereas large potato granules are ovoid and have a larger volume
compared to surface area 
[@vandeveldeStarchMorphology2002; @janeAnthologyStarch1994; @reichertStarchBible1913b, 364-365; @seidemannStarchAtlas1966, 174-176]. 
Another potentially important factor 
from our results is the size of the calculus deposit. We found a 
`r strength_cor` `r direct_cor` correlation between biofilm size and retained 
starch granules (Figure \@ref(fig:cor-plot)A);
a result that contradicts findings from archaeological contexts
[@wesolowskiEvaluatingMicrofossilContent2010; @dudgeonDietGeography2014].
When the concentration of starch granules
per mg calculus is considered, the correlation is weaker, but still present
(Figure \@ref(fig:cor-plot)B).
While the larger deposits contain a higher absolute count, our findings also suggest
that they contain a slightly higher concentration of starches. This may also explain
the lower mean retention of starch granules in mixed treatments compared to
wheat treatments. Wheat treatment samples (mean = `r subset(anova_tbl, treatment == "wheat")$mean` mg) were on average larger than mixed treatment samples (mean = `r subset(anova_tbl, treatment == "mix")$mean` mg) (Table \@ref(tab:anova-tbl)); and while mixed treatment solutions contained the highest mean overall
granule counts, wheat treatment samples had the highest mean starch retention.
Further research is needed to determine why this differs from previous archaeological
findings.

>>>>

We have also shown that the size of the starch granules influences the likelihood
of incorporation into the calculus. Starch granules larger than 20 $\mu$m in
maximum length were underrepresented in the calculus samples compared to the original 
starch solutions, an effect that was consistent across all three treatments. Medium
granules (10--20 $\mu$m) were often over-represented
(Table \@ref(tab:sample-prop-tbl), 
and 
Figure \@ref(fig:ratio-plots)).
Large potato granules were most affected, potentially because of the greater
size-range. They can reach up to 100 $\mu$m in maximum length, whereas wheat 
granules generally only reach up to 35 $\mu$m 
[@gismondiStarchGranulesData2019; @haslamDecompositionStarch2004; @seidemannStarchAtlas1966, 174-176].
Granule morphology may also play a role. Large wheat granules 
are lenticular and have a larger surface area compared 
to volume, whereas large potato granules are ovoid and have a larger volume
compared to surface area 
[@vandeveldeStarchMorphology2002; @janeAnthologyStarch1994; @reichertStarchBible1913b, 364-365; @seidemannStarchAtlas1966, 174-176]. 
Another potentially important factor 
from our results is the size of the calculus deposit. We found a 
`r strength_cor` `r direct_cor` correlation between size of biofilm deposit and
retained starch granules (Figure \@ref(fig:cor-plot)A), meaning larger calculus
deposits contain a higher quantity of granules;
a result that contradicts findings from archaeological contexts
[@wesolowskiEvaluatingMicrofossilContent2010; @dudgeonDietGeography2014].
When the concentration of starch granules
per mg calculus is considered, the correlation is weaker, but still present
(Figure \@ref(fig:cor-plot)B).
While the larger deposits contain a higher absolute count, our findings also suggest
that they contain a slightly higher concentration of starches. This may also explain
the lower mean retention of starch granules in mixed treatments compared to
wheat treatments. Wheat treatment samples (mean = `r subset(anova_tbl, treatment == "wheat")$mean` mg) were on average larger than mixed treatment samples (mean = `r subset(anova_tbl, treatment == "mix")$mean` mg) (Table \@ref(tab:anova-tbl)); and while mixed treatment solutions contained the highest mean overall
granule counts, wheat treatment samples had the highest mean starch retention.
Further research is needed to determine why this differs from previous archaeological
findings.

### 2022-05-31 analysis/paper/intro.Rmd
incorporate reviewer comments

<!-- clearly stating the general problem being addressed by this particular study -->
Despite this, our knowledge of dental calculus and the incorporation pathways of 
the various markers is limited [@radiniFoodPathways2017], as is our 
knowledge of information-loss caused by these pathways. Additionally, the methods
we use to extract and analyse dental calculus, and make inferences on past diets
represent another potential source of bias. Studies on both archaeological and
modern individuals have explored these biases in more detail.
Extraction methods were tested 
by Tromp and colleagues [-@trompEDTACalculus2017], specifically regarding
decalcification using HCl or EDTA. 
The authors found significantly more starches with the EDTA extraction method 
than the HCl extraction method; however, as noted by the authors, comparisons 
involving archaeological calculus are problematic due to variability between and 
within individuals. 
Studies conducted on modern humans [@leonardDentalCalculus2015] 
and non-human primates [@powerChimpCalculus2015], 
have explored how well microremains (phytoliths and starches)
extracted from dental calculus represent the actual dietary intake.
These studies are justifiably limited, 
despite meticulous documentation and observation, due to unknown variables and 
uncertainty involved in this kind of *in vivo* research. Dental calculus is a complex 
oral biofilm with a multifactorial aetiology and variable formation rates both 
within and between individuals [@jepsenCalculusRemoval2011; @haffajeeBiofilmPosition2009], 
contributing to
the stochasticity of starch representation being observed in numerous studies.
Additionally, the concentration of oral $\alpha$-amylase differs both between and
within individuals [@froehlichEffectOral1987; @naterHumanAmylase2005],
causing different rates of hydrolysis of the starch granules present in the oral
cavity. Add to this the effects of the many different methods
of starch processing, as well as post-depositional processes that are still being 
explored [@graneroStarchTaphonomy2020], and it becomes clear that using 
dental calculus to reconstruct diet is a highly unpredictable process.

>>>>

<!-- clearly stating the general problem being addressed by this particular study -->
Despite this, our knowledge of dental calculus and the incorporation pathways of 
the various markers is limited [@radiniFoodPathways2017], as is our 
knowledge of information-loss caused by these pathways. Additionally, the methods
we use to extract and analyse dental calculus, and make inferences on past diets
represent another potential source of bias. Studies on both archaeological and
modern individuals have explored these biases in more detail.
Extraction methods were tested 
by Tromp and colleagues [-@trompEDTACalculus2017], specifically regarding
decalcification using HCl or EDTA. 
The authors found significantly more starches with the EDTA extraction method 
than the HCl extraction method; however, as noted by the authors, comparisons 
involving archaeological calculus are problematic due to variability between and 
within individuals. 
Studies conducted on modern humans [@leonardDentalCalculus2015] 
and non-human primates [@powerChimpCalculus2015; @powerRepresentativenessDental2021] 
have explored how well microremains (phytoliths and starches)
extracted from dental calculus represent the actual dietary intake.
These studies are justifiably limited, 
despite meticulous documentation and observation, due to unknown variables and 
uncertainty involved in this kind of *in vivo* research. Dental calculus is a complex 
oral biofilm with a multifactorial aetiology and variable formation rates both 
within and between individuals [@jepsenCalculusRemoval2011; @haffajeeBiofilmPosition2009], 
contributing to
the stochasticity of starch representation being observed in numerous studies.
Additionally, the concentration of oral $\alpha$-amylase differs both between and
within individuals [@froehlichEffectOral1987; @naterHumanAmylase2005],
causing different rates of hydrolysis of the starch granules present in the oral
cavity. Add to this the effects of the many different methods
of starch processing [@hardyRecoveringInformation2018], as well as post-depositional
processes that are still being explored
[@graneroStarchTaphonomy2020; @mercaderExaggeratedExpectations2018],
and it becomes clear that using 
dental calculus to reconstruct diet is a highly unpredictable process.

### 2022-05-31 analysis/paper/paper.Rmd
incorporate reviewer comments

```{r microscope-fig, fig.cap="Microscope image of a biofilm sample exposed to the potato starch solution. Potato granules can be seen within a bacterial community. Scale bar = 20 $\\mu$m.", out.width="50%"}
knitr::include_graphics(here("analysis/figures/starches_w_bar.jpg"))
```

>>>>

```{r microscope-fig, fig.cap="Microscope image of a biofilm sample exposed to the potato starch solution. Potato granules can be seen within a bacterial community. Scale bar = 20 $\\mu$m.", fig.show='hold', out.width="40%"}
knitr::include_graphics(c(here("analysis/figures/starches_w_bar.jpg"), here("analysis/supplementary-materials/figures/SNAP-103412-0006.jpg")), )
```
